Article title

### 2017-07-31 paper-webserver.tex
FIX: verbing

% costs of bacterial infections
Foodborne ilnesses are increasingly difficult to track; ingrediants are globally sourced and contamination can occur at any point in the supply-chain.
The resulting consumer infections translate into actual costs: in Canada, an estimated 4000 hospitalizations and 105 deaths per year are thought to be the result of various pathogens \cite{thomas2015estimates}.
In the United States, 55961 hospitalizations and 1351 deaths per year \cite{scallan2011foodborne} are due to foodborne illnesses.
Futhermore, the associated economic costs have been known for decades \cite{todd1989costs}.
% surveillance
Public-health agencies routinely monitor emerging threats, of which \textit{Salmonella enterica} and \textit{Escherichia coli} are amongst the two most common \cite{kozak2013foodborne}.
In \textit{Salmonella enterica} \cite{bell2016recent}, \textit{Escherichia coli} \cite{fratamico2016advances}, and others pathogens \cite{ronholm2016navigating}, a variety of bioinformatic approaches to surveillance are now based on whole-genome sequencing (WGS).

% traditional reference lab techniques
To begin analyzing genomes, researchers designate particular strains of bacterial isolates using identifiers known as subtypes.
Traditional reference laboratory techniques handled subtyping through methods such as serotyping, pulsed-field gel electrophoresis (PFGE), or Sanger sequencing, amongst others \cite{ronholm2016navigating}.
PFGE is unique amongst the set; restriction enzymes are used to produce DNA fragments visible as specific banding patterns.
Public health agencies use PFGE to track outbreaks by comparing band patterns of isolates \cite{kozak2013foodborne}.
However, as WGS rarely produces complete circular genomes and, in PFGE, cutting by a restriction enzyme is dependent on protein structure, bioinformatic-analogs to PFGE are difficult to develop.
Unlike PFGE, both serotyping and MLST have WGS-based computational methods which mimick results of the wet-lab techniques.
Serotyping uses antibodies to detect the presence of specific cell surface antigens and multi-locus sequence typing (MLST), using Sanger sequencing, examines differences in housekeeping genes which are required to maintain cell function.
By comparing a pre-collected set of reference gene sequences against a WGS result, we can predict wet-lab serotyping results \cite{whiteside2016superphy}, Virulence Factor (VF) results \cite{joensen2014real}, and antimicrobial resistance (AMR) results \cite{mcarthur2013comprehensive}.
A similar approach can be taken for MLST \cite{larsen2012multilocus}.

% potential of WGS
WGS also expands on traditional methods: due to the inclusion of the entire genome, WGS-based methods offer better resolution and context for evaluating the spread of outbreaks \cite{ronholm2016navigating}.
In laboratory-based MLST, instead of examining selected single-nucleotide polymorphisms (SNPs) in a subset of housekeeping genes, newer WGS-based methods, such as core genome MLST (cgMLST), attempt to cover every common gene.
WGS approaches also expands on VF and AMR gene detection to cover every known sequence.
Furthermore, using WGS we can study the relations between genomes, changes in which VF or AMR genes are present over time, and mutations in genes \cite{ronholm2016navigating}.
Entire datasets and results can also be shared in real-time \cite{ronholm2016navigating}.

% Drawbacks - genetics
However, these WGS-based methods only examine the underlying genetics, not the expressed phenotype; the presence of a particular gene sequence does not necessitate expression.
Correspondingly, more target sequences are identified using WGS; if we take a tradition approach to making threat assessments by looking for known VFs or AMR genes, the large number of results make it difficult to guage risk \cite{ronholm2016navigating}.
The effectiveness of WGS-based methods also rely on the accuracy and breadth of reference datasets and agencies must weigh the cost of developing expertise in bioinformatic approaches, while laboratory based phenotypic tests remain inexpensive \cite{koser2012routine}.

% drawbacks - big data
Traditionaly, bioinformatics software was developed \textit{ad hoc} without employing software engineering principles \cite{de2015trends}.
Tools were often script-based, with custom data formats, and only suitable for small collections of data \cite{de2015trends}.
Since WGS-technology expanded the availability of data, systems engineering principles are important to handle code failures across the entire bioinformatics pipeline \cite{schatz2015biological}.
The field also lags behind other big-data industries: bioinformatics software still struggles to perform computations in real-time, stores data in inflexible databases \cite{de2015trends}, and lacks a common application programming interface (API) linking resources \cite{swaminathan2016review}.

@@

% why Superphy
To make bioinformatic approaches user-friendly for front-line clinicians and laboratories, a variety of web services have been developed in Denmark \cite{joensen2014real,larsen2012multilocus}, Taiwan \cite{liu2016construction}, and in collaborations \cite{hasman2015detection}.
While useful, arguably \cite{fricke2014bacterial}, the lack of bioinformatic standards and common infrastructure technologies make it difficult to fully leverage the big-data results from WGS.
Existing software also lack ways to integrate results from multiple genomes, and examine the connections between them.

% Superphy
Previously we developed Superphy \cite{whiteside2016superphy}, an online platform targeting \textit{E. coli} which determines Shiga toxin subtypes, AMR genes using CARD, and provides statistical significance tests between subgroups in associated metadata and pan-genome regions.
However, we struggled to provide real-time computing of results and to accomdate an ever-growing reference database.
Furthermore, due to the various dependencies involved, the platform was difficult for other researchers to independently replicate.

>>>>

% costs of bacterial infections
Tracking foodborne ilnesses is increasingly difficult; ingrediants are globally sourced and contamination can occur at any point in the supply-chain.
The resulting consumer infections translate into actual costs: foodborne illnesses cause an estimated 4000 hospitalizations and 105 deaths per year in Canada \cite{thomas2015estimates}, and 55961 hospitalizations and 1351 deaths per year in the United States \cite{scallan2011foodborne}.
Futhermore, many studies \cite{todd1989costs,scharff2010health} document the associated financial and productivity costs.
% surveillance
Public-health agencies routinely monitor emerging threats, of which \textit{Salmonella enterica} and \textit{Escherichia coli} are amongst the two most common \cite{kozak2013foodborne}.
In \textit{Salmonella enterica} \cite{bell2016recent}, \textit{Escherichia coli} \cite{fratamico2016advances}, and others pathogens \cite{ronholm2016navigating}, a variety of bioinformatic approaches to surveillance are now based on whole-genome sequencing (WGS).

% traditional reference lab techniques
To begin analyzing genomes, researchers designate particular strains of bacterial isolates using identifiers known as subtypes.
Traditional reference laboratory techniques handled subtyping through methods such as serotyping, pulsed-field gel electrophoresis (PFGE), or Sanger sequencing, amongst others \cite{ronholm2016navigating}.
PFGE is unique amongst the set; restriction enzymes are used to produce DNA fragments visible as specific banding patterns.
Public health agencies use PFGE to track outbreaks by comparing band patterns of isolates \cite{kozak2013foodborne}.
However, as WGS rarely produces complete circular genomes and, in PFGE, cutting by a restriction enzyme is dependent on protein structure, developing bioinformatic-analogs to PFGE is difficult.
Unlike PFGE, both serotyping and MLST have WGS-based computational methods which mimick results of the wet-lab techniques.
Serotyping uses antibodies to detect the presence of specific cell surface antigens and multi-locus sequence typing (MLST), using Sanger sequencing, examines differences in housekeeping genes which are required to maintain cell function.
By comparing a pre-collected set of reference gene sequences against a WGS result, we can predict wet-lab serotyping results \cite{whiteside2016superphy}, Virulence Factor (VF) results \cite{joensen2014real}, and antimicrobial resistance (AMR) results \cite{mcarthur2013comprehensive}.
MLST software uses a similar approach \cite{larsen2012multilocus}.

% potential of WGS
WGS also expands on traditional methods: due to the inclusion of the entire genome, WGS-based methods evaluate the spread of outbreaks with better resolution and context \cite{ronholm2016navigating}.
In laboratory-based MLST, instead of examining selected single-nucleotide polymorphisms (SNPs) in a subset of housekeeping genes, newer WGS-based methods, such as core genome MLST (cgMLST), attempt to cover every common gene.
WGS approaches also expands on VF and AMR gene detection to cover every known sequence.
Furthermore, using WGS we can study the relations between genomes, changes in which VF or AMR genes are present over time, and mutations in genes \cite{ronholm2016navigating}.
Entire datasets and results can also be shared in real-time \cite{ronholm2016navigating}.

% Drawbacks - genetics
However, these WGS-based methods only examine the underlying genetics, not the expressed phenotype; the presence of a particular gene sequence does not necessitate expression.
Correspondingly, more target sequences are identified using WGS; if we take a tradition approach to making threat assessments by looking for known VFs or AMR genes, guaging risk is difficult due to the large number of results \cite{ronholm2016navigating}.
The effectiveness of WGS-based methods also rely on the accuracy and breadth of reference datasets and agencies must weigh the cost of developing expertise in bioinformatic approaches, while laboratory based phenotypic tests remain inexpensive \cite{koser2012routine}.

% drawbacks - big data
Traditionaly, bioinformatics software was developed \textit{ad hoc} without employing software engineering principles \cite{de2015trends}.
Tools were often script-based, with custom data formats, and only suitable for small collections of data \cite{de2015trends}.
Since WGS-technology expanded the availability of data, bioinformatic pipelines are larger and more complicated.
We must apply systems engineering principles to handle code failures throughout \cite{schatz2015biological}.
The bioinformatics field also lags behind other big-data industries: software still struggles to perform computations in real-time, stores data in inflexible databases \cite{de2015trends}, and lacks a common application programming interface (API) linking resources \cite{swaminathan2016review}.

@@

% why Superphy
To make bioinformatic approaches user-friendly for front-line clinicians and laboratories, a variety of web services have been developed in Denmark \cite{joensen2014real,larsen2012multilocus}, Taiwan \cite{liu2016construction}, and in collaborations \cite{hasman2015detection}.
While useful, arguably \cite{fricke2014bacterial}, leveraging the big-data results from WGS is difficult due to the lack of bioinformatic standards and common infrastructure technologies.
Integrating results from multiple genomes, and examining shared connections, will be important for new software.

% Superphy
Previously we developed Superphy \cite{whiteside2016superphy}, an online platform targeting \textit{E. coli} which determines Shiga toxin subtypes, AMR genes using CARD, and provides statistical significance tests between subgroups in associated metadata and pan-genome regions.
However, we struggled to provide real-time computing of results and to accomdate an ever-growing reference database.
Furthermore, due to the various dependencies involved, independently replicating the platform was difficult for other researchers.
